% report.tex — Skeleton for a scientific report on Böðn's developmental foundations
%
% Compile with: ./build.sh (or: latexmk -pdf report.tex)
%
% This is a SKELETON — sections marked [TODO] need expansion with original writing.
% Citations are real and verifiable; verify DOIs before final submission.

\documentclass[12pt, a4paper]{article}

% --- Packages ---
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{geometry}
\geometry{margin=2.5cm}
\usepackage{setspace}
\onehalfspacing
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\hypersetup{colorlinks=true, linkcolor=blue, citecolor=blue, urlcolor=blue}
\usepackage{natbib}
\bibliographystyle{apalike}
\usepackage{enumitem}

% --- Project name ---
% \dh produces the eth character (ð) via T1 fontenc; \Bodn{} ensures
% the trailing space is not consumed after the macro.
\newcommand{\Bodn}{B\"o\dh{}n}

% --- Title ---
\title{Developmental Foundations of Tangible Interactive Learning Systems\\
       for Early Childhood:\\
       \large The \Bodn{} Project}
\author{
  Mathias Axell\\
  {\small Independent researcher}\\
  {\small \texttt{[email]}}
}
\date{\today}

% =============================================================================
\begin{document}
\maketitle

% --- Abstract ---
\begin{abstract}
\Bodn{} is an open-source, battery-powered tangible learning system designed for
children aged 3--7. The device combines physical controls (buttons, rotary
encoders, toggle switches) with multimodal feedback (LED animations, synthesized
audio, screen graphics) to deliver play experiences that target core executive
functions---working memory, inhibitory control, and cognitive flexibility---as
well as sensorimotor, cognitive, language, and social-emotional development.
This paper reviews the developmental science that informs \Bodn{}'s design,
maps each feature to established developmental domains, and identifies areas
for future development. We argue that tangible interfaces offer advantages over
touchscreen-only systems for early childhood learning, and that game-based,
guided play can meaningfully support executive function development during the
critical preschool window.

\medskip
\noindent\textbf{Keywords:} executive functions, tangible user interfaces,
early childhood, guided play, physical computing, open-source hardware
\end{abstract}

\tableofcontents
\newpage

% =============================================================================
\section{Introduction}
\label{sec:introduction}

The preschool years (ages 3--7) represent a critical period for the development
of executive functions (EF)---the cognitive control processes that enable
goal-directed behaviour \citep{diamond2013executive}. A large body of evidence
links early EF to school readiness \citep{blair2007relating}, social competence,
and lifelong well-being \citep{moffitt2011gradient}. Unlike IQ, which is
relatively stable, executive functions are highly malleable during early
childhood and respond to targeted practice
\citep{diamond2013executive, montroy2016development}.

At the same time, the proliferation of touchscreen devices has raised concerns
about passive screen time in young children \citep{radesky2015mobile}. The
World Health Organization recommends no more than one hour of sedentary screen
time per day for children aged 2--4 \citep{who2019guidelines}. This creates a
tension: digital systems can deliver adaptive, engaging practice that physical
toys cannot, yet the medium itself may undermine the developmental benefits.

Tangible user interfaces (TUI)---physical objects that bridge the digital and
physical worlds---offer a potential resolution. Since \citet{ishii1997tangible}
introduced the concept, a growing literature has explored TUI for educational
contexts, finding advantages in engagement, spatial understanding, and
embodied learning \citep{marshall2007tangible, price2003tangibles}.

This paper presents \textbf{\Bodn{}}, an open-source tangible learning system that
combines physical controls with adaptive, game-based experiences designed around
the developmental science of early childhood. We review the theoretical
foundations, describe the system design, map each feature to developmental
domains, and identify directions for future work.

% =============================================================================
\section{Background}
\label{sec:background}

% --- 2.1 Executive Functions ---
\subsection{Executive Functions in Early Childhood}
\label{sec:ef}

\citet{miyake2000unity} proposed the influential three-factor model of executive
functions: \emph{working memory} (updating), \emph{inhibitory control}
(inhibition), and \emph{cognitive flexibility} (shifting). While originally
validated in adults, subsequent work has confirmed this structure in preschoolers,
with important developmental nuances \citep{montroy2016development}.

\textbf{Working memory} is the ability to hold and manipulate information in
mind. In 4-year-olds, typical capacity is 2--4 items \citep{diamond2013executive}.
Tasks like remembering a sequence of coloured lights and reproducing it from
memory directly exercise this capacity.

\textbf{Inhibitory control} encompasses both response inhibition (suppressing a
prepotent motor response) and interference control (resisting distracting
information). The classic ``day--night'' task, where children must say ``night''
when shown a sun card, illustrates the challenge for preschoolers
\citep{diamond2013executive}.

\textbf{Cognitive flexibility} requires disengaging from one rule set and
engaging a new one. The Dimensional Change Card Sort (DCCS) task
\citep{zelazo2006dccs} is the standard assessment: children sort cards first by
colour, then by shape. Most 3-year-olds fail the switch; most 5-year-olds
succeed. The transition period (ages 3--5) represents a critical window for
practice.

Longitudinal data confirm that self-regulation (a broader construct overlapping
substantially with EF) improves steeply between ages 3 and 5, with individual
differences in developmental trajectory already predicting kindergarten
readiness \citep{montroy2016development}. The practical implication is that
\emph{when} intervention happens matters: the same dose of practice delivered at
age 3--5 is likely to produce larger gains than at age 7+.

The EF-training literature itself is mixed. \citet{diamond2016conclusions}
reviewed dozens of intervention studies and reached a sobering conclusion:
computerised working-memory training generalises poorly beyond the trained task,
while play-based, embodied, and socially embedded approaches (martial arts,
mindfulness, curricula such as Tools of the Mind) produce broader and more
durable transfer. The authors identify several common features of programs that
work --- sustained engagement, challenge calibrated to the child's current
level, multimodal practice, and low-stakes failure. \Bodn{}'s design is
explicitly shaped by this evidence: every game mode is playful rather than
drill-like, difficulty scales adaptively, feedback is multimodal, and errors are
gentle.

% --- 2.2 Guided Play ---
\subsection{Guided Play and Playful Learning}
\label{sec:guided_play}

\citet{weisberg2013guided} define guided play as adult-scaffolded activities
where children explore within a structured environment designed to make
discovery productive. Guided play occupies a middle ground between free play
(child-directed, no adult structure) and direct instruction (adult-directed,
no child agency). Evidence suggests that guided play produces deeper learning
than either extreme for concept acquisition and problem-solving
\citep{hirshpasek2015putting}.

\citet{skene2022guided} reported the most comprehensive quantitative evidence to
date: a systematic review and meta-analysis of 39 studies covering 3{,}800
children, finding that guidance during play produced greater gains than both
free play and direct instruction for early mathematics, shape knowledge, and
task-switching (an executive-function measure). The effect was strongest when
the child retained agency over how the activity unfolded even while an adult
(or an environment) set the frame. \citet{hassingerdas2017more} extend this
framework to well-designed games, arguing that games sit naturally on the
guided-play continuum: they supply clear goals and rules while preserving
voluntary engagement and child-led choice within those rules. \Bodn{}'s game
modes are deliberately structured this way --- each mode has a bounded rule
system (the ``frame''), but within the frame the child chooses when to play,
how fast to progress, and what to explore.

This framing also helps explain why open-ended modes like Mystery Box and Tone
Lab sit alongside rule-based modes like Simon and Rule Follow without
contradiction. Both are species of guided play: the environment's structure
differs, but in each case the child's action remains voluntary and
self-directed.

% --- 2.3 Tangible User Interfaces ---
\subsection{Tangible User Interfaces for Young Children}
\label{sec:tui}

\citet{ishii1997tangible} introduced the concept of tangible user interfaces
as a way to ``give physical form to digital information.'' For young children,
who are still developing fine motor skills and abstract thinking, the physical
affordances of tangible interfaces offer several advantages over touchscreens:

\begin{itemize}[noitemsep]
  \item \textbf{Embodied interaction}: Physical manipulation engages motor
    planning, proprioception, and spatial reasoning simultaneously
    \citep{price2003tangibles}.
  \item \textbf{Reduced cognitive load}: Physical affordances (a button looks
    pressable, a knob looks turnable) reduce the interface learning burden,
    freeing cognitive resources for content \citep{marshall2007tangible}.
  \item \textbf{Multimodal feedback}: Tactile confirmation (button click,
    encoder detent) combines with visual and auditory feedback to strengthen
    sensorimotor associations.
\end{itemize}

Empirical work specifically targeting preschool-age children supports these
claims. \citet{antle2013embodied} observed that 3--6-year-olds preferred
physical--digital interactions over screen-only equivalents and sustained
engagement longer; physicality, Antle argues, bridges concrete and abstract
thinking in a way screens alone cannot. \citet{manches2012tangibles} showed
that the deliberate, sequential nature of physical manipulation itself
exercises working memory more deeply than rapid button-tapping: searching for,
selecting, and placing a physical object is an extended motor--cognitive act
that offloads some intermediate state into the physical arrangement, freeing
mental resources for the task itself. \citet{sylla2012tinkrbook} reported that
tangible story tokens increased narrative complexity and sustained engagement
compared with screen-only storytelling in young children.

\citet{horn2012hybrid} make an important qualification: pure-tangible
interfaces are not automatically better than graphical ones. The clearest gains
come from hybrid designs that combine tangible input with rich on-screen (and
on-board) feedback. \Bodn{}'s architecture follows this hybrid pattern ---
every interaction is physical, every response is multimodal, and the two
screens are strictly used for feedback rather than as the interaction surface
themselves.

% --- 2.4 Multimodal Feedback and Sensory Integration ---
\subsection{Multimodal Feedback and Sensory Integration}
\label{sec:multimodal}

Young children benefit from redundant, multimodal feedback---receiving the same
information through multiple sensory channels simultaneously
\citep{price2003tangibles}. In \Bodn{}, every user action produces visual
(LED/screen), auditory (tone/sound), and tactile (physical click/detent)
feedback. This redundancy:

\begin{itemize}[noitemsep]
  \item Strengthens cross-modal neural connections
  \item Provides confirmation even if one channel is missed
  \item Supports different learning preferences and sensory profiles
\end{itemize}

The specifically cross-modal case is worth drawing out, because \Bodn{} exploits
it directly. \citet{maurer2006shape} replicated the classic Bouba/Kiki effect
in 2.5-year-olds, demonstrating that children form non-arbitrary
sound--shape associations well before literacy. \citet{dolscheid2014thickness}
showed that even four-month-old infants map higher pitch to higher spatial
position --- a prelinguistic cross-modal correspondence. These capacities are
the foundation for the Tone Lab's morphing blob: the shape responds to timbre
(round to spiky) and the vertical position responds to pitch (low to high),
giving the child a visual channel onto the same signal their ears are
processing. The goal is to make the sound's structure perceptible rather than
merely audible.

% --- 2.5 Music and Auditory Development ---
\subsection{Auditory Processing and Musical Exposure}
\label{sec:auditory}

Early exposure to structured sound environments supports auditory processing
skills that underpin language acquisition and musical development
\citep{kraus2010music, trainor2005critical}. Active sound production (as
opposed to passive listening) is particularly effective.

Two more specific lines of evidence inform \Bodn{}'s audio design. First,
\citet{gordon2003learning}'s Music Learning Theory argues that tonal and
rhythmic vocabulary should be built through open, in-tune exploration before
any formal notation is introduced --- children need to encounter many
well-formed musical patterns before they can produce or analyse them
themselves. The Tone Lab's strict-pentatonic constraint and the Sequencer's
fixed tempo grid both reflect this principle: the child explores freely but
everything they produce is musically well-formed. Second,
\citet{watson1985perception} and \citet{rochat2001infant} argue that the
perception of oneself as a \emph{cause} of sound is strongest when input and
audible effect are contingent within a narrow temporal window. \Bodn{}'s audio
path is therefore engineered to keep input-to-sound latency in the
low-millisecond range (the native C mixer runs on a dedicated CPU core),
preserving the tight contingency loop that underwrites early agency.

% --- 2.6 Bilingualism ---
\subsection{Bilingualism and Executive Function}
\label{sec:bilingualism}

A substantial literature documents enhanced executive functions in bilingual
children, attributed to the constant need to manage two active language systems
\citep{bialystok2011reshaping}. The bilingual advantage is most pronounced for
inhibitory control and cognitive flexibility---the same skills that \Bodn{}'s
game modes target.

The bilingual advantage has been contested:
\citet{paap2015bilingualism} argued that positive results reflect publication
bias rather than a robust effect, and \citet{lehtonen2018bilingualism}'s
meta-analysis of adult studies reached a similar conclusion once publication
bias was modelled. However, in the specific context of early childhood (ages
3--5), where language systems are still being established, the evidence for
enhanced inhibitory control remains more consistent
\citep{bialystok2011reshaping, barac2014cognitive}. The pragmatic conclusion
is that dual-language exposure is worth pursuing for its intrinsic
communicative value --- being able to understand and speak two languages ---
even if the downstream executive-function benefits are smaller or more
context-dependent than early work suggested. Sweden provides a particularly
favourable setting for early bilingualism: English-language media exposure is
pervasive, and Swedish preschools increasingly introduce English informally.

\subsubsection*{Letter Case and Early Literacy}

Research on letter-name knowledge shows that preschoolers recognise uppercase
letters before lowercase ones---uppercase knowledge predicts lowercase
acquisition with a facilitation factor exceeding 16$\times$
\citep{piasta2014letter}. However, since more than 95\% of printed text uses
lowercase forms, prolonged exposure to only uppercase delays the transition to
reading real text \citep{treiman2006case}. \Bodn{} therefore uses \textbf{title
case} (first letter uppercase, remainder lowercase) for all child-facing
labels, giving the child the familiar uppercase anchor while building lowercase
familiarity.

\subsubsection*{Dual-Language Labeling}

\citet{byersheinlein2013bilingualism} demonstrated that bilingual infants
distinguish their two languages from birth using rhythmic and phonological cues,
countering the common concern that simultaneous exposure causes confusion.
\citet{tan2024translation} showed that \emph{translation equivalents}---knowing
a word in one language---significantly increases the probability of learning the
corresponding word in the other language, especially at younger ages. A peer-reviewed synthesis of
dual-language-learner research concludes that vocabulary acquisition in both
languages is supported when children receive consistent, paired exposure to
the same concepts in each language \citep{barac2014cognitive}. \Bodn{}'s printed NFC cards therefore present both
Swedish and English labels simultaneously (e.g.\ ``Katt / Cat''), while the
on-device UI shows one language at a time (switchable in settings) to respect
the display's space constraints.

% --- 2.7 Pretend Play ---
\subsection{Pretend Play and Social-Cognitive Development}
\label{sec:pretend_play}

\citet{lillard2013impact} reviewed the evidence linking pretend play to
cognitive development, finding associations with theory of mind, narrative
skills, and emotional understanding. While the causal direction remains debated,
the correlational evidence is strong, and pretend play is universally observed
across cultures during the preschool years.

Two \Bodn{} modes are explicitly designed as scaffolded pretend-play
environments. \emph{Spaceship Cockpit} casts the child as the captain of a
small ship, with the AI companion ``Stellar'' announcing scenarios (``Asteroids
ahead! Steer out of the way!'') via TTS. The physical encoders, toggles, and
arcade buttons are reframed as throttle, shields, and steering --- the same
hardware controls become different fictional objects depending on the scenario.
This kind of symbolic substitution (treating object A \emph{as if} it were
object B) is a hallmark of preschool pretend play and is thought to contribute
to theory-of-mind development by requiring the child to hold two
interpretations of the same object in mind simultaneously
\citep{lillard2013impact}. \emph{Story Mode} complements this with branching
narrative: the child guides a character through a story world by choosing
between spoken options at decision points. The narrative frame supplies the
pretend context; the branching structure ensures the child's choices matter.
Blippa adds a third, looser register --- because the interaction is goal-free,
children spontaneously treat the reader as a supermarket scanner or transit
gate, producing entirely child-initiated pretend scenarios that the device
simply supports.

% --- 2.8 Growth Mindset and Failure Tolerance ---
\subsection{Feedback, Failure, and Growth Mindset}
\label{sec:mindset}

\citet{dweck2006mindset} distinguishes between fixed mindset (ability is innate)
and growth mindset (ability develops through effort). The type of feedback
children receive after failure shapes which mindset they develop. Process-focused
feedback (``You tried a different strategy, good!'') promotes growth mindset;
outcome-focused feedback (``Wrong!'') promotes fixed mindset.

\Bodn{}'s feedback design follows this principle: errors produce soft, warm cues
(gentle tone, dim red glow) rather than punitive signals. The intent is to
frame failure as information, not judgement.

Several concrete design decisions follow from this stance. Success feedback is
not louder or more dramatic than error feedback --- it is simply different in
tone. This is deliberate: overly celebratory success animations implicitly
punish failure by contrast. No mode uses a red ``wrong'' buzzer, a loudly
descending ``sad trombone,'' or a shaking screen; every error is acknowledged
with a soft chime and a warm, dim visual cue and then immediately gives the
child another try. Simon, for example, allows up to three attempts per
sequence and only grows the sequence length on success, so the child is never
confronted with an impossibly long sequence. Rule Follow switches rules after
several successes rather than after errors, so the game becomes harder as the
child improves rather than punishing failure with added difficulty. High-Five
Friends auto-restarts after three misses rather than forcing a session-ending
screen. These decisions collectively aim to keep the emotional stakes low
enough that the child continues engaging --- a necessary condition for the
kind of repeated, voluntary practice that \citet{diamond2016conclusions}
identify as the core ingredient of effective EF interventions.

% =============================================================================
\section{System Design}
\label{sec:system}

% --- 3.1 Hardware ---
\subsection{Hardware Architecture}
\label{sec:hardware}

\Bodn{} is built around an ESP32-S3 microcontroller with the following physical
interface components:

\begin{itemize}[noitemsep]
  \item 8 mini push buttons (colour-coded, via MCP23017 I2C expander)
  \item 5 arcade buttons with PWM-controlled LEDs (via PCA9685)
  \item 2 toggle switches (via MCP23017)
  \item 3 rotary encoders with push buttons (navigation, volume, throttle)
  \item 2.8'' ILI9341 TFT primary display (240$\times$320, landscape)
  \item 1.8'' ST7735 TFT secondary display (128$\times$160, portrait)
  \item 2$\times$ WS2812 8-LED sticks + 92-LED perimeter strip (RGB)
  \item MAX98357A I2S amplifier with 3W speaker
  \item INMP441 I2S MEMS microphone
  \item 6600\,mAh LiPo battery with thermal monitoring (DS18B20$\times$2)
  \item PN532 NFC reader (I2C) for tangible card interaction
\end{itemize}

A complete bill of materials, pin assignments, and thermal-protection
escalation table is maintained in the project's \texttt{docs/hardware.md}.
The device enclosure is a wooden box chosen for its warmth and durability;
the goal is for the object to feel like a toy rather than a piece of
consumer electronics.

% --- 3.2 Software Architecture ---
\subsection{Software Architecture}
\label{sec:software}

The firmware is written in MicroPython and follows a modular architecture:

\begin{itemize}[noitemsep]
  \item \textbf{Rule engines} (\texttt{*\_rules.py}): Pure logic modules with
    no hardware dependencies, testable via standard pytest on the host.
  \item \textbf{Screen modules} (\texttt{ui/*.py}): Render game state to
    displays and LEDs; handle input events.
  \item \textbf{Audio engine}: 16-voice native C mixer running on a dedicated
    CPU core, with priority channels (music, SFX pool, UI feedback) and
    built-in tone generators (sine, square, sawtooth, noise).
  \item \textbf{Session manager}: Enforces play time limits, breaks, and
    quiet hours.
  \item \textbf{i18n system}: All user-facing strings localised
    (Swedish default, English available).
  \item \textbf{Native C modules}: performance-critical paths (audio mixer,
    display DMA, bitmap blit, MCP23017 input scan, NeoPixel animation
    engine, Game of Life step kernel) are implemented as MicroPython user
    C modules. Each has a pure-Python fallback so the design can be
    iterated in the simulator before the custom firmware is built.
  \item \textbf{Asset resolver}: a single \texttt{resolve()} function routes
    all sound and TTS lookups, checking the SD card first and falling back
    to flash. This lets hand-recorded narration transparently override
    generated TTS, and keeps firmware size independent of the media
    library.
\end{itemize}

The emphasis on pure-logic rule engines reflects a deliberate testing
strategy: every game mode's rules can be exercised under CPython on a
laptop with \texttt{pytest}, while hardware-dependent code is tested
separately via the Wokwi simulator or directly on the device. This keeps
the developmental logic reviewable by non-embedded collaborators and
dramatically shortens the iteration loop.

% =============================================================================
\section{Developmental Analysis}
\label{sec:analysis}

This section maps each \Bodn{} feature to established developmental domains and
analyses coverage. Table~\ref{tab:matrix} presents the full feature--domain
matrix.

% --- 4.1 Feature Mapping ---
\subsection{Feature--Domain Mapping}
\label{sec:mapping}

Each subsection below describes one of the fourteen currently shipped game
modes, identifies the primary developmental aspects it targets, and links the
design decisions back to the literature reviewed in Section~\ref{sec:background}.

\subsubsection{Simon (H\"armapa)}
The Simon memory game directly targets \textbf{working memory} capacity. The
starting sequence length of 2 items is calibrated to the lower bound of
4-year-old WM capacity \citep{diamond2013executive}. Each successful round
adds one item, providing adaptive difficulty within the child's zone of
proximal development \citep{vygotsky1978mind}.

\subsubsection{Rule Follow (Regelleken)}
Modelled after the DCCS task \citep{zelazo2006dccs}, Rule Follow alternates
between ``match'' and ``opposite'' rules. The ``opposite'' rule requires
\textbf{inhibitory control} (suppressing the prepotent matching response), while
the rule switch exercises \textbf{cognitive flexibility}. The frequency of rule
switches (every 4--6 correct answers) is within the range used in DCCS research.

\subsubsection{Mystery Box (Mysterium)}
An open-ended \textbf{cause-and-effect} exploration mode with no failure state.
The design follows the ``guided play'' model \citep{weisberg2013guided}: the
environment is structured (specific button--colour mappings, discoverable
combinations) but the child's exploration is entirely self-directed.

\subsubsection{Fl\"ode}
A \textbf{spatial reasoning} and \textbf{problem-solving} puzzle that requires
aligning gaps in vertical walls. The snap-to-grid mechanic reduces motor
precision demands while preserving the cognitive challenge.

\subsubsection{Garden of Life (Tr\"adg\aa rden)}
A \textbf{pattern recognition} mode based on Conway's Game of Life. The child
plants coloured seeds and observes emergent behaviour---an implicit introduction
to complex systems and self-organisation. The mode exercises no executive
functions but provides rich visual stimulation and cause-and-effect learning.

\subsubsection{Soundboard (Ljudbr\"ada)}
An \textbf{auditory processing} and \textbf{open exploration} mode.
Customisable sound banks allow parents to create themed experiences. The volume
encoder provides fine motor practice with immediate auditory feedback.

\subsubsection{Spaceship Cockpit (Rymdskepp)}
The most developmentally comprehensive mode, targeting \textbf{cognitive
flexibility} (5 scenario types requiring different inputs),
\textbf{listening comprehension} (TTS-driven instructions), and
\textbf{imaginative play} (structured pretend-play as a spaceship captain).
The adaptive difficulty system adjusts timeout windows based on the child's
recent performance, maintaining an appropriate challenge level.

\subsubsection{Story Mode (Saga)}
A data-driven branching narrative engine where the child listens to narrated
scenes and makes choices via arcade buttons to steer the plot. Primary targets:
\textbf{listening comprehension} (must listen to narration to choose),
\textbf{bilingual exposure} (full narration in both Swedish and English), and
\textbf{imaginative play} (guiding a character through a narrative world).
Stories are defined as script files on the SD card; adding new stories requires
no code changes.  The first story adapts Beatrix Potter's \emph{The Tale of
Peter Rabbit} (public domain) into a choose-your-own-adventure with four endings.

\subsubsection{High-Five Friends (Heja-kompisar)}
A reaction-time game targeting \textbf{hand-eye coordination} and
\textbf{inhibitory control}. Animals ``pop up'' on illuminated arcade buttons
and the child must tap the lit button before it disappears. The reaction window
shrinks from 2\,s to 0.5\,s as the child succeeds, providing adaptive
difficulty within the child's zone of proximal development
\citep{vygotsky1978mind}. Gentle miss feedback (soft sound, dim animation) and
automatic restart after three misses support \textbf{persistence} while keeping
emotional stakes low \citep{dweck2006mindset}. The mode also exercises
\textbf{auditory processing} (distinct hit/miss sounds) and
\textbf{self-regulation} (managing the frustration of near-misses).

\subsubsection{Sortera --- NFC Classification Game}
A tangible implementation of the Dimensional Change Card Sort (DCCS) task
\citep{zelazo2006dccs}, targeting \textbf{cognitive flexibility} and
\textbf{inhibitory control}. The device announces a sorting rule (``Find the
animals!''), and the child scans NFC cards. After several correct matches the
rule switches to a different dimension (``Find all the red ones!''), requiring
the child to inhibit the previous sorting strategy. Sortera is the first mode
to provide \textbf{active bilingual reinforcement}: on each correct scan, the
screen displays both the Swedish and English word (e.g.\ ``Katt / Cat'')
matching the physical card, and bilingual TTS announces both words---primary
language first, then secondary \citep{tan2024translation}. This dual-label
approach leverages the translation-equivalent bootstrapping effect while keeping
game instructions in the configured language for clarity.

\subsubsection{Räkna --- Progressive Math with Tangible Numbers}
A six-level NFC math mode targeting \textbf{pattern recognition} (subitising),
\textbf{sequential thinking} (counting and equation order),
\textbf{problem solving} (composing equations step by step), and
\textbf{fine motor} coordination (handling multiple cards per equation). Räkna
implements Bruner's Concrete--Pictorial--Abstract progression
\citep{bruner1966instruction,willingham2017manipulatives}: a physical dot-pattern
card is placed on the reader (concrete), the screen renders matching dots
(pictorial), and from Level~4 onwards the numeral and operator cards are
introduced (abstract). Dice-face dot patterns at low levels rather than
numerals reflect the developmental evidence that subitising is the foundation
of number sense, not numeral recognition
\citep{clementssarama2014learning}. Levels span free exploration (ages
3--4), quantity matching and more/less comparison (4--5), addition and
subtraction with tokens (5--6), and symbolic equations (6--7+). A left-to-right
linear number path on the display implements Siegler and Ramani's
\citeyearpar{siegler2009playing} finding that board-game-style number lines
measurably improve preschool number competence. Bilingual TTS announces each
quantity in Swedish and English, carrying the translation-equivalent
bootstrapping effect \citep{tan2024translation} into a cognitively engaging
math task. The mode is always self-paced and never timed --- a deliberate
design choice given the evidence that timed math exercises produce anxiety
even in preschoolers. Together with Sortera, Räkna addresses the early-math
domain that \citet{duncan2007school} identified as the single strongest
predictor of later academic achievement.

\subsubsection{Blippa --- Free-Play NFC Discovery}
A goal-less, failure-less companion to the other NFC modes targeting
\textbf{open exploration}, \textbf{cause and effect}, and \textbf{imaginative
play}. Tapping any owned card---from any card set the device knows about---
produces a full-screen emoji, a bilingual label, and either a procedural POS
``blip'' or an optional per-card sample. There is no rule to follow, no score,
and no wrong scan: the mode exists specifically to turn a growing collection
of physical cards into a low-friction exploration toy. This design is
motivated by evidence that children's causal learning is strongest when
exploration is self-directed rather than instructed \citep{bonawitz2011double},
and it intentionally creates a surface for the cashier and transit-gate
pretend play that 3--5-year-olds gravitate to, linking open exploration with
imaginative play. Unknown card identifiers within a known card set produce a
deterministic ``mystery blip'' whose pitch is derived from a stable hash of
the id, so a given card always sounds the same---reinforcing object
permanence and rehearsal of cause-effect contingencies.

\subsubsection{Sequencer (Sequenser)}
A loop-based step sequencer targeting \textbf{working memory},
\textbf{sequential thinking}, \textbf{pattern recognition}, and
\textbf{creative expression}. Five percussion tracks (arcade buttons) and one
melody track (eight mini buttons mapping to a pentatonic scale) are recorded
into an 8- or 16-step grid during looped playback. Toggle switches control
play/pause, step count, and metronome---making abstract musical concepts
tangible. The mode introduces computational thinking through creative play
\citep{papert1980mindstorms, bers2018coding}: building patterns, running them,
hearing something unexpected, and debugging by toggling individual steps. BPM is adjustable via
encoder (70--140), providing fine motor practice with immediate auditory
feedback.

\subsubsection{Tone Lab (Ljudlabb)}
A real-time synthesiser sandbox targeting \textbf{auditory processing},
\textbf{cause-and-effect learning}, \textbf{visual processing via cross-modal
binding}, and \textbf{creative expression}. Two encoders control pitch (strict
pentatonic over two octaves) and timbre (a five-step morph from smooth sine to
fuzzy noise). Five arcade buttons snap to fixed pentatonic pitches so nothing
ever sounds ``wrong''---consistent with Gordon's Music Learning Theory
\citep{gordon2003learning}, which argues tonal vocabulary is built through
open but in-tune exploration. Eight mini buttons provide hold-to-apply
effects (vibrato, tremolo, pitch bend up/down, octave jump up/down, stutter,
harmony); each effect is a single reversible gesture, keeping the
cause-and-effect link transparent \citep{watson1985perception}. All effects
run inside the native audio mixer, so every input change is reflected in the
waveform within milliseconds---preserving the tight contingency loop that
young children need to perceive themselves as the cause of the sound
\citep{rochat2001infant}.

Two visual channels onto the same signal reinforce cross-modal learning. A
live oscilloscope shows the actual waveform, and a central blob morphs from
round (sine) to spiky (noise) in lockstep with the timbre encoder---
exploiting the Bouba/Kiki correspondence first documented by
\citet{kohler1929gestalt} and replicated in toddlers by
\citet{maurer2006shape}. The blob's vertical position tracks pitch (higher
tones higher on screen), drawing on the pitch-height correspondence present
even in infants \citep{dolscheid2014thickness}. A toggle switch swaps the
big oscilloscope between the primary and secondary displays so older
children or accompanying adults can inspect the waveform directly while the
child remains focused on the expressive blob.

% --- 4.2 Coverage Matrix ---
\subsection{Coverage Matrix}
\label{sec:coverage}

\begin{table}[htbp]
\centering
\caption{Feature--developmental domain coverage matrix. %
  $\bullet\bullet\bullet$ = primary target;
  $\bullet\bullet$ = significant;
  $\bullet$ = incidental;
  blank = not applicable.}
\label{tab:matrix}
\small
\begin{tabular}{lcccccccccccccc}
\toprule
\textbf{Domain} & \rotatebox{70}{Simon} & \rotatebox{70}{Rule Follow}
  & \rotatebox{70}{Mystery Box} & \rotatebox{70}{Fl\"ode}
  & \rotatebox{70}{Garden} & \rotatebox{70}{Soundboard}
  & \rotatebox{70}{Spaceship} & \rotatebox{70}{Story Mode}
  & \rotatebox{70}{High-Five} & \rotatebox{70}{Sequencer}
  & \rotatebox{70}{Tone Lab}
  & \rotatebox{70}{Sortera} & \rotatebox{70}{R\"akna}
  & \rotatebox{70}{Blippa} \\
\midrule
Working memory       & $\bullet\bullet\bullet$ & $\bullet\bullet$ & & $\bullet$ & & & $\bullet\bullet$ & $\bullet\bullet$ & $\bullet$ & $\bullet\bullet\bullet$ & $\bullet$ & $\bullet$ & $\bullet\bullet$ & \\
Inhibitory control   & $\bullet$ & $\bullet\bullet\bullet$ & & & & & $\bullet\bullet$ & & $\bullet\bullet$ & & & $\bullet\bullet$ & $\bullet$ & \\
Cognitive flexibility & & $\bullet\bullet\bullet$ & & & & & $\bullet\bullet\bullet$ & & & $\bullet\bullet$ & $\bullet$ & $\bullet\bullet\bullet$ & $\bullet$ & \\
\midrule
Cause \& effect      & & & $\bullet\bullet\bullet$ & & $\bullet\bullet$ & $\bullet\bullet$ & & $\bullet\bullet$ & $\bullet\bullet$ & $\bullet\bullet$ & $\bullet\bullet\bullet$ & $\bullet\bullet$ & $\bullet\bullet$ & $\bullet\bullet\bullet$ \\
Pattern recognition  & $\bullet\bullet$ & & $\bullet\bullet$ & & $\bullet\bullet\bullet$ & & & & & $\bullet\bullet\bullet$ & $\bullet$ & $\bullet\bullet\bullet$ & $\bullet\bullet\bullet$ & $\bullet$ \\
Spatial reasoning    & & & & $\bullet\bullet\bullet$ & & & & & & & & & $\bullet$ & \\
Sequential thinking  & $\bullet\bullet\bullet$ & & & $\bullet\bullet$ & & & & $\bullet\bullet$ & & $\bullet\bullet\bullet$ & & & $\bullet\bullet\bullet$ & \\
Problem solving      & & & & $\bullet\bullet\bullet$ & & & $\bullet$ & & & $\bullet\bullet$ & & & $\bullet\bullet$ & \\
\midrule
Fine motor           & $\bullet$ & $\bullet$ & $\bullet$ & $\bullet\bullet$ & $\bullet$ & $\bullet$ & $\bullet\bullet$ & $\bullet$ & $\bullet$ & $\bullet\bullet$ & $\bullet\bullet$ & $\bullet\bullet$ & $\bullet\bullet$ & $\bullet\bullet$ \\
Hand--eye coord.     & $\bullet\bullet$ & $\bullet\bullet$ & & $\bullet\bullet$ & & & $\bullet\bullet$ & & $\bullet\bullet\bullet$ & $\bullet\bullet$ & $\bullet$ & & $\bullet$ & $\bullet$ \\
Auditory processing  & $\bullet\bullet$ & & & & & $\bullet\bullet\bullet$ & $\bullet\bullet$ & $\bullet\bullet\bullet$ & $\bullet\bullet$ & $\bullet\bullet$ & $\bullet\bullet\bullet$ & $\bullet\bullet$ & $\bullet\bullet$ & $\bullet\bullet$ \\
Visual processing    & $\bullet\bullet$ & $\bullet\bullet$ & $\bullet\bullet\bullet$ & $\bullet\bullet$ & $\bullet\bullet\bullet$ & & $\bullet\bullet$ & $\bullet\bullet$ & $\bullet\bullet$ & $\bullet\bullet$ & $\bullet\bullet$ & $\bullet\bullet$ & $\bullet\bullet\bullet$ & $\bullet\bullet$ \\
\midrule
Listening comp.      & & & & & & $\bullet$ & $\bullet\bullet\bullet$ & $\bullet\bullet\bullet$ & & & & $\bullet$ & $\bullet\bullet$ & $\bullet$ \\
Bilingual exposure   & $\bullet$ & $\bullet$ & & $\bullet$ & $\bullet$ & $\bullet$ & $\bullet$ & $\bullet\bullet\bullet$ & $\bullet$ & $\bullet$ & $\bullet$ & $\bullet\bullet$ & $\bullet\bullet$ & $\bullet\bullet$ \\
\midrule
Self-regulation      & $\bullet\bullet$ & $\bullet\bullet\bullet$ & & & $\bullet$ & & $\bullet\bullet$ & $\bullet$ & $\bullet\bullet$ & & & $\bullet\bullet$ & $\bullet$ & \\
Persistence          & $\bullet\bullet$ & $\bullet$ & & $\bullet\bullet\bullet$ & & & $\bullet$ & $\bullet$ & $\bullet\bullet$ & $\bullet\bullet$ & & $\bullet$ & $\bullet\bullet$ & \\
Imaginative play     & & & & & & & $\bullet\bullet\bullet$ & $\bullet\bullet\bullet$ & & $\bullet\bullet$ & $\bullet$ & & & $\bullet\bullet$ \\
\midrule
Open exploration     & & & $\bullet\bullet\bullet$ & & $\bullet\bullet\bullet$ & $\bullet\bullet\bullet$ & & & & & $\bullet\bullet\bullet$ & $\bullet$ & $\bullet$ & $\bullet\bullet\bullet$ \\
Divergent thinking   & & & $\bullet\bullet\bullet$ & & $\bullet\bullet$ & & & & & $\bullet\bullet\bullet$ & $\bullet\bullet$ & & $\bullet$ & $\bullet$ \\
Creative expression  & & & & & $\bullet\bullet$ & $\bullet\bullet\bullet$ & & & & $\bullet\bullet\bullet$ & $\bullet\bullet\bullet$ & & & \\
\bottomrule
\end{tabular}
\end{table}

% --- 4.3 Gap Analysis ---
\subsection{Gap Analysis}
\label{sec:gaps}

The current fourteen-mode catalogue yields strong coverage (three or more
features at primary or significant level) across every developmental aspect
except spatial reasoning. The shift is the result of several deliberate
additions that each filled a specific gap identified in earlier revisions of
this matrix. Story Mode gave listening comprehension, bilingual exposure, and
imaginative play a second primary-level feature apiece. The Sequencer
introduced creative expression and computational thinking. High-Five Friends
added dedicated hand--eye coordination with adaptive timing. The three NFC
modes then added a tangible manipulation layer: Sortera provides a tangible
Dimensional Change Card Sort \citep{zelazo2006dccs} strengthening cognitive
flexibility and inhibitory control; R\"akna opens the early-mathematics
domain that \citet{duncan2007school} identified as the single strongest
predictor of later academic achievement, following Bruner's
Concrete--Pictorial--Abstract progression \citep{bruner1966instruction} and
Siegler and Ramani's \citeyearpar{siegler2009playing} linear-number-board
evidence; and Blippa closes the tangible cause-and-effect loop without any
goal or failure state, aligning with \citet{bonawitz2011double} on
exploration-driven causal learning. Finally, Tone Lab added a third
primary-level auditory-processing mode alongside Soundboard and Story Mode
and---uniquely---targeted cross-modal sound--shape perception
\citep{kohler1929gestalt, maurer2006shape, dolscheid2014thickness}, a
developmental capacity no previous mode addressed directly.

The single remaining under-covered domain is:

\begin{enumerate}[noitemsep]
  \item \textbf{Spatial reasoning}: still primarily Fl\"ode, with R\"akna's
    left-to-right number path contributing only incidentally. A tangram or
    construction mode would add breadth, and the predictive link between
    preschool spatial assembly and later mathematics
    \citep{verdine2014deconstructing} makes this a high-leverage gap to
    close.
\end{enumerate}

Two further areas remain thinner than the rest even though they pass the
three-feature threshold: problem solving is presently Fl\"ode (primary),
Sequencer (debugging), and R\"akna (equation building), so a scavenger-hunt
NFC puzzle mode would usefully broaden it; and listening comprehension rests
on only two primary features (Spaceship, Story Mode), so a dedicated
language-switching game could both strengthen it and further leverage
bilingual exposure for EF development \citep{bialystok2011reshaping}.

% =============================================================================
\section{Discussion}
\label{sec:discussion}

\subsection{Tangible Interfaces vs.\ Touchscreen Applications}

The dominant form factor for children's digital learning tools is the
tablet-based app. \Bodn{}'s commitment to physical controls is a deliberate
departure, grounded in three specific arguments.

First, the embodied-cognition literature holds that thinking in young children
is scaffolded by the body: the actions a child can perform constrain and shape
the concepts they can form \citep{antle2013embodied, price2003tangibles}. A
button that clicks, an encoder that detents, and a toggle that latches give the
child proprioceptive feedback independent of the screen --- the hand
\emph{knows} something happened even before the eyes register it. Touchscreen
taps provide none of this. Second, physical controls act as an external memory
\citep{manches2012tangibles}: an NFC card on the table is visible in the
peripheral field whether or not the child is currently attending to it, which
supports the kind of extended planning behaviour that a rapidly scrolling
touchscreen actively disrupts. Third, and pragmatically, physical devices are
easier to regulate: a wooden box with a power switch is harder to use
compulsively than an app embedded in a general-purpose device that also
contains YouTube.

Commercial tablet-based EF training programs (Cogmed and similar) have produced
mixed results in preschool populations, with transfer effects narrow to the
trained tasks \citep{diamond2016conclusions}. \Bodn{} does not attempt to
produce drill-based training; it instead embeds EF demands inside games the
child chooses to play repeatedly, in line with the broader-transfer
approaches Diamond \& Ling identified as effective.

\subsection{Adaptive Difficulty and the Zone of Proximal Development}

\citet{vygotsky1978mind} proposed that learning is most efficient when a child
is supported to operate just beyond what they can already do unaided --- the
Zone of Proximal Development (ZPD). Several \Bodn{} modes implement explicit
adaptive-difficulty systems that aim to keep the child in this zone. Simon
grows sequence length by one item per success and never shortens it unless the
child fails repeatedly, so the demand tracks the child's capacity. Spaceship
shortens the response timeout as the child succeeds; a child who is struggling
receives more time, a child who is mastering the scenarios receives less.
High-Five Friends shrinks the reaction window from 2\,s down to 0.5\,s across a
round. R\"akna's six-level progression makes the same principle explicit at a
coarser granularity: the child begins with free dot-card exploration and moves
to symbolic equations over months or years as readiness emerges.

An implicit assumption of the ZPD framing is that the \emph{right} level of
challenge is knowable in advance. In practice, \Bodn{}'s adaptive loops use
simple rules based on recent success/failure counts rather than any formal
model of the child's skill. This is adequate for single-session adaptation but
does not persist across sessions. A Bayesian skill-estimation layer is listed
in Section~\ref{sec:future} as a plausible extension.

\subsection{Feedback Design and Growth Mindset}

The growth-mindset literature \citep{dweck2006mindset} argues that feedback
focused on process (``you tried a different strategy --- good'') produces more
persistent learners than feedback focused on outcome or innate ability. \Bodn{}
operationalises this at the level of every error sound and every animation:
mistakes produce soft, warm cues; successes produce differently-shaped cues
rather than proportionally larger ones; and no mode ever displays a
``game over'' screen in large red letters. This is an unusually strong
stance --- many commercial children's games use loud failure feedback
explicitly because it provokes the ``one more try'' response --- and the
trade-off is accepted deliberately: \Bodn{} prioritises emotional safety over
short-term engagement loops, betting that lower-frustration play sustains
longer-term engagement \citep{diamond2016conclusions}.

A subtler design decision concerns achievements and progression markers.
\Bodn{} has no stars, no points, no leaderboards, and no unlockable content
gated behind performance. The rationale is two-fold: first, extrinsic rewards
can undermine intrinsic motivation in young children; second, rewards create
winners and losers in a single-device household, which is undesirable. Level
progression in R\"akna, for example, is surfaced only in the parental
controls, not to the child.

\subsection{Session Limits and Healthy Technology Use}

The WHO \citep{who2019guidelines} recommends no more than one hour of sedentary
screen time per day for children aged 2--4. The key word is \emph{sedentary}:
the concern is not the photons emitted by a screen but the passive consumption
pattern that typically accompanies them. \Bodn{} challenges the category: its
screens exist, but they respond only to physical action, and the child cannot
progress without moving. This is closer to ``physical play with visual
feedback'' than to traditional screen time, and the session-limit framework is
sized accordingly.

\Bodn{}'s parental controls allow configurable session duration, daily session
count, mandatory breaks between sessions, and quiet hours. Session-end is
signalled by a gradual wind-down animation rather than an abrupt cutoff ---
the child learns that ``time to stop'' is a predictable, normal part of play,
rather than a punishment. All play-history data stays on the device; no
telemetry is uploaded to a cloud, and the parental web dashboard is served
directly from the ESP32 over the local network.

\subsection{Limitations}

Several limitations should be acknowledged. First, no formal user studies or
controlled evaluations have been conducted; the claims in this paper are
design-rationale arguments grounded in the developmental literature, not
experimental findings about \Bodn{} itself. Second, \Bodn{} is currently a
single-user device: every mode assumes one child at a time, and cooperative
or turn-taking features are not yet implemented. This limits the system's
ability to support socio-cognitive development in peer contexts. Third, the
bilingual-advantage literature that motivates some of \Bodn{}'s design
decisions is actively contested
\citep{paap2015bilingualism, lehtonen2018bilingualism}; the dual-language
features should be viewed primarily as supporting communicative competence in
both languages, with downstream EF effects as a possible but uncertain
bonus. Fourth, hardware constraints (low-resolution displays, no haptic
feedback beyond the physical controls, a modest 3\,W speaker) limit the
range of sensory experiences the device can deliver. Finally, the author is
not a developmental psychologist; the literature review draws on published
work but has not been peer-reviewed by domain experts.

% =============================================================================
\section{Future Work}
\label{sec:future}

\begin{enumerate}
  \item \textbf{NFC tangible interaction layer}: Integrate a PN532 NFC reader to
    enable tagged physical objects (cards, stickers, figurines) as input.
    Research on tangible user interfaces consistently shows that physical
    manipulation benefits learning in preschoolers through reduced cognitive load,
    embodied engagement, and support for classification tasks
    \citep{marshall2007tangible, antle2013embodied, manches2012tangibles}. Planned
    NFC modes include:
    \begin{itemize}[noitemsep]
      \item \emph{Sortera} --- a tangible classification game implementing the DCCS
        task \citep{zelazo2006dccs} with physical cards, targeting cognitive
        flexibility and inhibitory control.
      \item \emph{Saga Builder} --- tangible storytelling where scanned
        character/setting cards trigger narration, extending the existing Story
        Mode with a richer physical input method
        \citep{sylla2012tinkrbook}.
      \item \emph{Vocabulary Explorer} --- bilingual word cards for active
        language switching, leveraging the bilingual EF advantage
        \citep{bialystok2011reshaping}.
      \item \emph{Memory Match} --- a physical search-and-select task that adds
        motor planning to working memory challenges
        \citep{manches2012tangibles}.
      \item \emph{R\"akna} --- a progressive math mode using dot-pattern NFC
        cards as tangible number tokens. Follows Bruner's
        Concrete-Pictorial-Abstract progression
        \citep{bruner1966instruction}: physical tokens (concrete)
        $\to$ dot displays (pictorial) $\to$ numerals (abstract). Six
        levels spanning subitising, counting, comparison, addition, and
        subtraction (ages 3--7). Grounded in evidence that early math
        is the strongest predictor of later academic achievement
        \citep{duncan2007school}, that linear number board games
        produce measurable and persistent gains in preschoolers'
        number competence \citep{siegler2009playing}, and that
        kindergarten number sense predicts math achievement through
        third grade \citep{jordan2009early}.
    \end{itemize}
    The NFC interaction pattern (find object $\to$ bring to reader $\to$ hold)
    gives children physical agency in a way that fixed buttons cannot, and has
    been commercially validated by products like Toniebox and Yoto Player for
    children as young as 3. See the companion document
    \texttt{nfc\_tangible\_learning.md} for the full research summary.

  \item \textbf{Empirical validation}: Conduct pre--post studies comparing
    EF measures (DCCS, digit span) in children using \Bodn{} vs.\ control
    activities.
  \item \textbf{Record \& Replay mode}: Implement voice recording and playback
    with effects (pitch shift, echo) to strengthen creative expression,
    auditory processing, and bilingual exposure.
  \item \textbf{Cooperative play}: Add turn-taking and collaborative modes to
    support social-cognitive development.
  \item \textbf{Adaptive difficulty modelling}: Implement Bayesian difficulty
    estimation to maintain each child in their zone of proximal development
    across play sessions.
  \item \textbf{Longitudinal tracking}: Use the existing session logging
    infrastructure to track developmental progress over months and years.
  \item \textbf{Active bilingual features}: Develop language-switching games
    that leverage the bilingual EF advantage.
\end{enumerate}

% =============================================================================
\section{Conclusion}
\label{sec:conclusion}

We have argued that tangible, open-source hardware can deliver developmentally
informed play experiences for young children that neither screen-based apps
nor traditional toys match alone. \Bodn{}'s design grounds each feature in
established developmental science: executive functions are exercised by Simon,
Rule Follow, Sortera, and Sequencer; early mathematics by R\"akna; bilingual
exposure by Story Mode, Sortera, R\"akna, and Blippa; creative expression by
Tone Lab and Sequencer; pretend play by Spaceship and Story Mode. Across all
fourteen modes a small set of recurring design principles do the heavy lifting
--- physical-first interaction with multimodal feedback, guided play with
preserved child agency, low-stakes failure, and contingent input--output loops
tight enough to support the perception of oneself as a cause.

The coverage analysis (Section~\ref{sec:coverage}) indicates that the current
catalogue addresses nearly every developmental aspect identified as relevant
for ages 3--7, with spatial reasoning as the one remaining gap. Future work
should close that gap, add empirical validation studies, and extend the
platform to cooperative play. Because \Bodn{} is fully open source --- hardware
designs, firmware, documentation, and the research behind the design decisions
--- we hope the approach can serve both as a platform for empirical work in
child development and as an invitation to other makers, educators, and
developmental researchers to collaborate on physical computing for young
children.

% =============================================================================
% References
\bibliography{report}

\end{document}
